Functional class scoring-based enrichment tests of Infinium DNA methylation data

### 2023-10-03 manuscript/manuscript.Rmd
writing starts

@@

>>>>

### Implementation overview

Functional enrichment analysis is a process of data summarisation from genes to gene sets (pathways).
This is made more complicated for Infinium array data due to the presence of multiple probes per gene,
meaning the data needs to be summarised from probes to genes to gene sets.

We propose three different approaches for applying enrichment analysis depending on when the differential
methylation step is conducted (**Figure 1**).

1. <strong>Differential methylation analysis is conducted on probes and these results are aggregated
to the gene and gene set level (called "LA").<.strong>
In this approach differential analysis was conduced on probe values using limma.
Differential probe methylation results (t-statistics) were then aggregated to the gene level by median.
These median t-statistics were subject to downstream enrichment analysis using a one-sample two-sided
t-test (a self-contained test) to estimate whether the t-statistic distribution of the gene set varies
from zero.
The use of a self-contained test is warranted as global hyper- or hypo-methylation is common between
conditions, where competitive tests may give misleading results.
The use of parametric tests is also warranted, as probe t-statistics typically exhibit a normal
distribution, even in the presence of global hyper- or hypo-methylation.

2. <strong>Differential methylation analysis conducted after probe-level methylation is aggregated to
gene-level (called "AL").</strong>
As each gene may have many probes, these individual data points were summarised into a single
value for each individual.
We used the mean of M-values for each probe belonging to a gene, which generated a new gene-level
M-value matrix.
This gene-level methylation matrix underwent differential analysis using limma, followed by an
enrichment test as above (one-sample two-tailed t-test).

3. <strong>Methylation values are aggregated to the gene set level for differential testing
(called "AA").</strong>
M-values for each probe were aggregated to the gene level by mean, and then aggregated again to the gene
set level by mean or median.
The resulting matrix of aggregated M-values underwent differential analysis with limma and the results
including log fold changes and significance values were used to infer pathway regulation.
This approach has a caveat in that it may partly violate the assumption of independent sampling, as
it is known that some genes are represented in multiple gene sets.

### Method validation

As there are no established methods for benchmarking Infinium array based pathway regulation using
known ground truth synthetic data, we used a split-and-analyse approach with a real dataset to
evaluate the three approaches outlined above.
We downloaded raw intensity EPIC IDAT files from GEO for study GSE158422 which consists of
lung tumour and normal adjacent tissues from 37 patients.
Probe annotations were obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor
package.
Reactome gene sets were downloaded from the MSigDB website.
From this set of samples, 18 patients were randomly selected into analysis A, where control and tumour
samples were compared.
From the remaining 19 samples, 18 were randomly selected into analysis B, where control and tumour
samples were compared as above to act as an independent validation of analysis A.
Analyses A and B were conducted using one of the three enrichment approaches outlined above, and the
gene sets meeting a 5% FDR threshold in the hyper- and hypo-methylated directions were saved.
Gene sets identified to be common in A and B analyses (FDR<0.05 and direction) were considered validated,
gene sets identified in only one method were not considered validated, and gene sets that gave discordant
direction (hyper- vs hypo-methylation) between A and B were classed as contradictory.
This process was repeated 50 times to rigourously evaluate method performance.

### Sensitivity analysis

The same dataset including all 37 patients underwent analysis using the LA method, gsameth and ebGSEA.
The effect size (enrichment score) of significantly differentially methylated pathways (FDR<0.05) was
recorded.
Then, a random subset of n patients was selected followed by pathway enrichment analysis with the
respective method.
The subset results were compared to the full n=37 results to test the sensitivity of these enrichment
methods at differenent effect size levels.
This process was repeated 20 times.

### Association of methylation with gene expression

As methylation at promoters and gene bodies is considered to have differing association with gene
expression, we conducted LA analysis at gene promoters, gene bodies and whole gene separately and
examined the relationship with RNA-seq based pathway regulation.
RNA-seq counts were downloaded from the GEO database (Accession: GSE158420).
HGNChelper was used to repair transmogrified gene names, followed by DESeq2 based differential analysis.
Enrichment analysis of RNA-seq data was conducted using the mitch package with REACTOME gene sets.

### Pathway-level differential methylation in assisted reproductive technology
(i) the effect of assisted reproductive technology DNA methylation

### Pathway-level differential methylation in hyperhomocysteinemia
(2) the association of elevated plasma homocysteine with methylation patterns.

@@

### Evaluation of FCS methods for Infinium methylation array data

We applied the three distinct approaches ("LA", "AL" and "AA") for pathway level differential
methylation analysis to a dataset consisting of lung tumour and adjacent normal tissue from 37 patients.
The dataset was split randomly into two groups of 18 patients and the overlap between subset was used to
assess the consistency and validity of findings with 50 simulations.
From the 1550 Reactome gene sets in the dataset, the AL method recorded the highest number of
"valid" pathways with an average of 919.8 being identified in both subsets, which was much higher than
LA and AA methods with 295.3 and 276.7 gene sets respectively (Figure 1A).
The AL method also recorded the lowest number of uncommon sets (382.0) which was better than the LA and
AA methods with 634.8 and 908.3 respectively (Figure 1B).
We also noticed that the LA method sometimes yielded contradictory results, so we quantified the number
of discordant observations, that is statistically significant but in the opposite direction.
True to this observation, the LA method gave 461.6 discordant pathways which indicates a high rate of
false-positives.
Meanwhile the AL and AA methods gave a mean of 0.1 and 0.02 discordant pathways respectively.
Taken together, these results indicate that the AL method gives the highest rate of common results
suggesting good accuracy in terms of sensitivity and specificity in this context.

### Comparision of AL method to existing pathway enrichment approaches
